# (PART) MAPPING FILES {-}

# Preparing Mapping Files
The mapping files are require to direct the pipeline where to look for the files containing the sequencing data.

- The format of the mapping files for `mothur` and `QIIME2` pipelines is slightly different. 
- We will demonstrate how to prepare both.
- Each mapping file will contain only the files that are parsed by bioinformatics analysis. 

> We will select files containing <65,000 sequences for demo purposes. We will start by getting the statistics of the samples.

## Explore read statistics 
- Are the files compressed?
- You may want to save space by compressing the unzipped files. 
- Navigate to the folder containing the fastq files and compress them using `gzip` function.

```bash
gzip *.fastq
```

> From this point foward we will assume that all the fastq files are in `fastq.gz` format.

- Using `seqkit stat` function to compute read statistics.
- We will store the output in `data/stats1/seqkit_stats.txt`.
- The seqkit output will contain useful information for preparing the mapping files.

```bash
mkdir -p data
mkdir -p data/stats1  
seqkit stat *.fastq.gz >data/stats1/seqkit_stats.txt

```

## Mothur mapping file
The mapping files for use with `mothur` pipeline can be generated automatically if using the whole dataset. However, if you are interested in just a fraction of the files, you can create it manually and specify what to include.


```{r}
library(tidyverse)
read_table("data/stats1/seqkit_stats.txt") %>% 
  mutate(sample_id = str_replace_all(file, ".*/", ""), .before=file) %>% 
  mutate(sample_id = str_replace_all(sample_id, "_\\d?.fastq.gz", "")) %>%
  filter(str_detect(file, "_1" )) %>% 
  mutate(file2 =file, .after = file) %>% 
  mutate(file2 = str_replace_all(file, "_1.fastq.gz", "_2.fastq.gz")) %>%
  distinct() %>% 
  group_by(sample_id) %>%
  filter(num_seqs <65000) %>% 
  arrange(num_seqs) %>% 
  ungroup() %>%
  select(sample_id, forward = file, reverse = file2) %>% 
  write_tsv("data/mothur_mapping_file.tsv")
  # saveRDS("RDataRDS/mothur_mapping_file.rds")

read_tsv("data/mothur_mapping_file.tsv") %>% 
  as.data.frame()

```

## Metadata for Mothur pipeline
- Mothur pipeline expects the design (metadata) file to have column headers. 
- We will extract only the desired number of samples.
- The first column header should be **group**.
- For more detail click [here](https://mothur.org/wiki/miseq_sop/).

```{r}
read_tsv("data/mothur_mapping_file.tsv") %>% 
  select(sample_id) %>% 
  inner_join(., readRDS("RDataRDS/metadata.rds"), by = "sample_id") %>% 
  select(group = sample_id, isolate) %>%
  write_tsv("data/mothur_metadata.tsv")
  # saveRDS("RDataRDS/mothur_metadata.rds")

read_tsv("data/mothur_metadata.tsv") %>% 
  as.data.frame()
```

## QIIME2 mapping file

```{r}
library(tidyverse)
read_table("data/stats1/seqkit_stats.txt") %>% 
  mutate(sample_id = str_replace_all(file, ".*/", ""), .before=file) %>% 
  mutate(sample_id = str_replace_all(sample_id, "_\\d?.fastq.gz", "")) %>%
  filter(str_detect(file, "_1" )) %>% 
  mutate(file2 =file, .after = file) %>% 
  mutate(file2 = str_replace_all(file, "_1.fastq.gz", "_2.fastq.gz")) %>%
  distinct() %>% 
  group_by(sample_id) %>%
  filter(num_seqs <65000) %>% 
  arrange(num_seqs) %>%
  ungroup() %>% 
  mutate(file = str_replace_all(file, "SRR", "$PWD/SRR")) %>%
  mutate(file2 = str_replace_all(file2, "SRR", "$PWD/SRR")) %>%
  select("sample-id" = sample_id, "forward-absolute-filepath" = file, "reverse-absolute-filepath" = file2) %>% 
  write_tsv("data/q2-pe-33-manifest.tsv")
  # saveRDS("RDataRDS/q2-pe-33-manifest.rds")
  
read_tsv("data/q2-pe-33-manifest.tsv") %>% 
  as.data.frame()
```

## Metadata for QIIME2 pipeline
- The TSV format is recommended for QIIME2 metadata. For more details click [here](https://docs.qiime2.org/2022.2/tutorials/metadata/).
- The first column is sample identifier i.e. `sample-id` NOT with underscore like `sample_id`.
- The command below will replace `sample_id` in the metadata to `sample-id`.

```{r}
read_tsv("data/q2-pe-33-manifest.tsv") %>% 
  select(sample_id = "sample-id") %>% 
  inner_join(., readRDS("RDataRDS/metadata.rds"), by = "sample_id") %>% 
  rename("sample-id" = sample_id) %>%
  write_tsv("data/q2-metadata.tsv")
  # saveRDS("RDataRDS/q2-metadata.rds")

read_tsv("data/q2-metadata.tsv") %>% 
  as.data.frame()
```



